HO unification from object language to meta language

### 2024-02-19 main.tex
wip intro

The object logic we are interested in is Dependent Type Theory (DTT),
for which we aim to implement a unification procedure \U{} using the ML
Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
Elpi comes equipped with the equational theory \Ue{}, comprising
$\eta\beta$ equivalence
and higher order unification restricted to the pattern
fragment~\cite{miller92jsc}.
We want \U{} to feature the same equational theory as \Ue{} but on the
object logic DTT.
Unfortunately the natural encoding of DTT in Elpi, which we refer to as \Fo{},
``underuses'' \Ue{} by restricting it to first-order unification problems only. 
To address this issue, we propose a better-behaved encoding, \Ho{},
demonstrate how to map unification problems in \Fo{}
to related problems in \Ho{}, and illustrate
how to map back the unifiers found by \Ue{}, effectively implementing
 \U{} on top of \Ue{} for the natural encoding \Fo{}.

@@

related~\cite{felty93lics}

>>>>

The object logic we are interested in is Dependent Type Theory (DTT),
for which we aim to implement a unification procedure \U{} using the ML
Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
Elpi comes equipped with the equational theory \Ue{}, comprising
$\eta\beta$ equivalence
and higher order unification restricted to the pattern
fragment~\cite{miller92jsc}.
We want \U{} to feature the same equational theory as \Ue{} but on the
object logic DTT.
Unfortunately the natural encoding of DTT in Elpi, which we refer to as \Fo{},
\marginpar{natural is not the right word}
``underuses'' \Ue{} by restricting it to first-order unification problems only. 
To address this issue, we propose a better-behaved encoding, \Ho{},
demonstrate how to map unification problems in \Fo{}
to related problems in \Ho{}, and illustrate
how to map back the unifiers found by \Ue{}, effectively implementing
 \U{} on top of \Ue{} for the natural encoding \Fo{}.

@@

Specifying and implementing a logic from scratch requires significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways: 1) variable binding and substitution are simplified when ML binders
represent object logic ones; 2) proof construction, and even proof search, is
greatly simplified by leveraging the unification procedure provided by the ML.
Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
which have been utilized to implement various formal systems such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873} and
even Higher Order Logic~\cite{books/sp/NipkowPW02}.

The object logic we are interested in is Dependent Type Theory (DTT),
and we want to code a type-class~\cite{wadler89}
solver for Coq~\cite{Coq-refman}.
Type-class solvers are proof search procedures based on back-chaining designated
lemmas, providing essential automation to widely used 
Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
constitute our test bed.

For example, these 3 designated lemmas (Instances in the TC slang)
are supposed to be animated, backward chained, by the TC solver.
fin n is the type of numbers smaller than n, nfact the number of prime
factors and Finite/Decision properties designated as classes what are
automatically proved by Coq using the designated instances.

\begin{coqcode}
Instance fin_fin n : Finite (fin n).
Instance nfact_dec n nf : Decision (nfact n nf).
Instance forall_dec A P : Finite A ~$\to$~ ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x, P x).
\end{coqcode}

for example here

\begin{coqcode}
  Check _ : Decision (forall y: fin 7, nfact y 3).
\end{coqcode}

A natural encoding (others are discussed section xx)
uses c for constants, all for forall, and then app and lam
for the lambda calculus part.

\begin{elpicode}
finite (app[c"fin", N]).
decision (app [c"nfact",N,NF]).
decision (all A x\ app[P, x]) :- (pi x\ decision (app[P, x])), finite A.
\end{elpicode}

unfortunately this does not work since the last rule cannt
backchain on the encoded goal

\begin{elpicode}
decision (all (app[c"fin",c"7"]) y\ app[c"nfact", y, c"3"])
\end{elpicode}

indeed this unif problem has no solution

\begin{elpicode}
app[c"nfact", y, c"3"] = app[P, y]
\end{elpicode}

in this paper we study a more sophisticated encoding

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm A P, (pi x\ decision (app[P, x])), finite A.
\end{elpicode}

This time Pm is HO for elpi, an the extra link is there to explain how to
get P out of it.

\begin{elpicode}
app[c"nfact", y, c"3"] = Pm y
Pm = y\app[c"nfact", y, c"3"]
P = lam (app[c"fin",c"7"]) a\ app[c"nfact", a, c"3"]
\end{elpicode}

We also see that this is not trivial, because in the rec call
we have now a beta redex.

\subsection{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%

our choice of encoding of DTT may look weird to the reader familiar with
LF, since used a shallow encoding of classes and binders, but not of the
``lambda calculus'' part of DTT. Here a more lightweight encoding
that unfortunately does not fit our use case

\begin{elpicode}
finite (fin N).
decision (nfact N NF).
decision (all A x\ P x) :- (pi x\ decision (P x)), finite A.
\end{elpicode}

but in DTT this is not always possible and not handy in our use case,
since the arity of constants is not fixed.

\begin{coqcode}
Fixpoint narr T n := if n is S m then T -> narr T m else T.
Definition nsum n : narr nat (n+1).
Check nsum 2   8 9 : nat.
Check nsum 3 7 8 9 : nat.
\end{coqcode}

moreover we use the same encoding for meta programming, or even just to provide
hand written rules. We want to access the syntax of OL, so our embedding cannot
be that shallow. We want to keep it shallow for the binders, but we need
a c, app and lam nodes.

another alternative

\begin{elpicode}
decision X :- unif X (all A x\ app[P, x]), (pi x\ decision (app[P, x])), finite A.
\end{elpicode}

gives up all half of what the ML gives us. moreover even if unif here embodies
he eq theory of DTT which is much stringer than the on of ML, we don't need 
it. according to our experience eta beta suffice, but HO is needed.

Note that this~\cite{felty93lics} is related and make the
discrepancy between the types of ML and DTT visible. In this case
one needs 4 application nodes. Moreover the objective is an encoding
of terms, proofs, not proof search. Also note the conv predicate,
akin to the unif we rule out.

This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.
